Porting the Steel Bank Common Lisp Compiler and Runtime to the Nintendo Switch NX Platform

### 2025-04-10 els2025-sbcl-nx/paper.tex
Minor fixups

\section{Relocation}\label{relocation}
Lisp objects contain absolute pointers, so that when the in-process state of all Lisp memory spaces are dumped to disk, the resulting \textit{core image} is filled with absolute pointers. Traditionally, image-based Lisp implementations such as SBCL map their address spaces to fixed addresses for two reasons: First, reloading an image into a process is easier this way, since objects can be mapped back in the new image directly to where they were in the process that dumped the image. Second, some architecture backends optimize code generation by hardwiring the address of certain Lisp objects directly into the machine code.

\subsection{Address Space Layout Randomization}
Because the NX operating system enforces address space layout randomization, the Lisp runtime cannot rely on being able to map its address spaces to fixed addresses. Hence, the runtime must relocate all objects present in the dumped image to the address spaces placed randomly by the operating system. As part of our work, we have extended the existing support for heap relocation to allow all other Lisp spaces used by SBCL to be relocated as well, with the result that the runtime can fix up all absolute pointers mapped in from the image to point to the random addresses given to the runtime by the operating system. In addition, we modified the Lisp code generation to produce strictly position-independent code.

However, code objects still present a special problem on the NX for this relocation process. A code object in SBCL is represented like other Lisp objects in-memory, in that it has a header word describing the type and size of code object. The code object also contains slots pointing to the debug information associated with the code object as well as other boxed objects which represent constants referenced by the code instructions. This is so that the garbage collector can scan the code object and fix up these boxed pointers as it does with other Lisp objects. Finally, these slots are followed by the raw machine instructions and function entry points representing the rest compiled Lisp code. The code object associated with the function defined in listing~\ref{lst:function} is illustrated in figure~\ref{fig:before-shrinkwrap}.

@@

\begin{figure}[h]
  \label{fig:before-shrinkwrap}
  \centering
  \includesvg[width=\linewidth]{Code object before shrinkwrapping.drawio.svg}
  \caption{Representation of the compiled code object (in boldface)
    for the function \protect\texttt{\#'f} in memory, along with the
    cons and symbol objects it references.}
\end{figure}

However, executable pages are not writable on the NX, so that the pointers to boxed Lisp data for the debug information and code constants cannot be fixed up as part of the relocation process on start-up. Furthermore, only the system loader is able to allocate executable pages by loading code from on-disk ELF .text sections, so that it isn't possible to first fix up boxed object pointers in the code object \emph{before} marking the code pages executable. In either case, the garbage collector would not be able to fix up these pointers either. Therefore, the code and data need to be separated, at first so that the runtime can relocate the absolute pointers to the Lisp objects, and later so that the moving garbage collector is able to fix up those references. This restriction requires the shrinkwrapping procedure to separate the data from the code and to rewrite the code in the image offline so that any references in code to code constants\footnote{As for the per-code-object slot containing debug information, we modified the Lisp system so that debug information is not accessed through the code object at all but rather through a weak hash table keyed by code object.} are into a dislocated read/write space instead of into the code object itself, which is on an executable page and hence not writable on the NX. To demonstrate this process in more detail, our modified offline shrinkwrapping process produces two artifacts from a normal Lisp image:

\begin{itemize}
\item \code{pie-shrinkwrap-sbcl.s}: A textual assembly file is generated with all code objects in the original core image listed under \code{.text} sections. The pointers to debug information and code constants inside the code object are dumped out as zeroes, and instead, the code constants referenced via the ARMv8 \code{LDR} instruction are displaced into a r/w \code{.data} section shared by all code objects. The \code{LDR} instructions are rewritten to access the constants from this section instead.
\item \code{pie-shrinkwrap-sbcl.o}: A binary object file with the rest of the Lisp data from the core image section is generated. The majority of the Lisp image is thus `shrinkwrapped' in large r/w \code{.data} section.
\end{itemize}

The reason for producing both a textual and binary artifact is that it is easier to rewrite and produce textual assembly for code, and it is faster to process and link a binary artifact for the rest of the data. Figure~\ref{fig:after-shrinkwrap} illustrates how the code object for the function \code{#'f} gets split up. The shrinkwrapping process then links both artifacts with the rest of the SBCL runtime to produce the final executable. The upshot of this process is that both the relocation process and the garbage collector can now scan and fix up all boxed pointers in the image at runtime without running into the read-only constraint for executable pages.

\begin{figure}[h]
  \label{fig:after-shrinkwrap}
  \centering
  \includesvg[width=\linewidth]{Code objects in memory after shrinkwrapping.drawio.svg}
  \caption{Representation of the same code object and the cons and
    symbol objects it references after code/data segregation and the
    rest of the shrinkwrapping process.}
\end{figure}

\subsection{Linkage between Lisp and foreign code}
Another part of the system impacted by the need for relocation is linkage between Lisp and foreign code. SBCL uses a linkage table that indirects calls to foreign code from Lisp so that foreign code can be loaded by the operating system at any address on startup and resolved in the linkage table. However, for callbacks into Lisp from foreign code, Lisp must create function pointers to snippets of assembly for callback entry allocated in static (i.e. never moved by GC) Lisp space. These pointers can then be passed to and called from foreign code. To ensure that such function pointers in a saved image are valid even after the Lisp spaces containing the callback entry points are relocated, the function pointers must somehow be fixed up so they are valid after startup. We managed to solve this problem by having the offline shrinkwrapping process move the callback entry code into another \code{.text} ELF section while recording fixup information, so that the operating system can load the callback entry code into executable pages, whereby the Lisp runtime at startup can rewrite the static function pointers to point to where the code was loaded using the recorded fixup information.

>>>>

\section{Relocation}\label{relocation}
Lisp objects contain absolute pointers, meaning when the in-process state of all Lisp memory spaces is dumped to disk, the resulting \textit{core image} is filled with absolute pointers. Traditionally, image-based Lisp implementations such as SBCL map their address spaces to fixed addresses for two reasons: First, reloading an image into a process is easier this way, since objects can be mapped back in the new image directly to where they were in the process that dumped the image. Second, some architecture backends optimize code generation by hardwiring the address of certain Lisp objects directly into the machine code.

\subsection{Address Space Layout Randomization}
Because the NX operating system enforces address space layout randomization, the Lisp runtime cannot rely on being able to map its address spaces to fixed addresses. Hence, the runtime must relocate all objects present in the dumped image to the address spaces placed randomly by the operating system. As part of our work, we have extended the existing support for heap relocation to allow all other Lisp spaces used by SBCL to be relocated as well, with the result being that the runtime can fix up all absolute pointers mapped in from the image to point to the random addresses given to the runtime by the operating system. In addition, we modified the Lisp code generation to produce strictly position-independent code.

However, code objects still present a special problem on the NX for this relocation process. A code object in SBCL is represented like other Lisp objects in-memory, in that it has a header word describing the type and size of code object. The code object also contains slots pointing to the debug information associated with the code object as well as other boxed objects which represent constants referenced by the code instructions. This is so that the garbage collector can scan the code object and fix up these boxed pointers as it does with other Lisp objects. Finally, these slots are followed by the raw machine instructions and function entry points representing the compiled Lisp code. The code object associated with the function defined in \autoref{lst:function} is illustrated in \autoref{fig:before-shrinkwrap}.

@@

\begin{figure}[h]
  \centering
  \includesvg[width=\linewidth]{Code object before shrinkwrapping.drawio.svg}
  \caption{Representation of the compiled code object (in boldface)
    for the function \protect\texttt{\#'f} in memory, along with the
    cons and symbol objects it references.}
  \label{fig:before-shrinkwrap}
\end{figure}

However, executable pages are not writable on the NX, meaning the pointers to boxed Lisp data for the debug information and code constants cannot be fixed up as part of the relocation process on start-up. Furthermore, only the system loader is able to allocate executable pages by loading code from on-disk ELF .text sections, meaning it isn't possible to first fix up boxed object pointers in the code object \emph{before} marking the code pages executable. In either case, the garbage collector would not be able to fix up these pointers either. Therefore, the code and data need to be separated, at first so that the runtime can relocate the absolute pointers to the Lisp objects, and later so that the moving garbage collector is able to fix up those references. This restriction requires the shrinkwrapping procedure to separate the data from the code and to rewrite the code in the image offline so that any references in code to code constants\footnote{As for the per-code-object slot containing debug information, we modified the Lisp system so that debug information is not accessed through the code object at all but rather through a weak hash table keyed by code object.} are into a dislocated read/write space instead of into the code object itself, which is on an executable page and hence not writable on the NX. To demonstrate this process in more detail, our modified offline shrinkwrapping process produces two artefacts from a normal Lisp image:

\begin{figure}[h]
  \centering
  \includesvg[width=\linewidth]{Code objects in memory after shrinkwrapping.drawio.svg}
  \caption{Representation of the same code object and the cons and
    symbol objects it references after code/data segregation and the
    rest of the shrinkwrapping process.}
  \label{fig:after-shrinkwrap}
\end{figure}

\begin{itemize}
\item \code{pie-shrinkwrap-sbcl.s}: A textual assembly file is generated with all code objects in the original core image listed under \code{.text} sections. The pointers to debug information and code constants inside the code object are zeroed out, and the code constants referenced via the ARMv8 \code{LDR} instruction are displaced into a r/w \code{.data} section shared by all code objects. The \code{LDR} instructions are rewritten to access the constants from this section instead.
\item \code{pie-shrinkwrap-sbcl.o}: A binary object file with the rest of the Lisp data from the core image section is generated. The majority of the Lisp image is thus `shrinkwrapped' into a large r/w \code{.data} section.
\end{itemize}

The reason for producing both a textual and a binary artefact is that it is easier to rewrite and produce textual assembly for code, and it is faster to process and link a binary artefact for the rest of the data. \autoref{fig:after-shrinkwrap} illustrates how the code object for the function \code{#'f} gets split up. The shrinkwrapping process then links both artefacts with the rest of the SBCL runtime to produce the final executable. The upshot of this process is that both the relocation process and the garbage collector can now scan and fix up all boxed pointers in the image at runtime without running into the read-only constraint for executable pages.

\subsection{Linkage between Lisp and foreign code}
Another part of the system impacted by the need for relocation is linkage between Lisp and foreign code. SBCL uses a linkage table that indirects calls to foreign code from Lisp so that foreign code can be loaded by the operating system at any address on startup and resolved in the linkage table. However, for callbacks into Lisp from foreign code, Lisp must create function pointers to snippets of assembly allocated in static (i.e. never moved by GC) Lisp space. These pointers can then be passed to and called from foreign code. To ensure that such function pointers in a saved image are valid even after the Lisp spaces containing the callback entry points are relocated, the function pointers must somehow be fixed up. We managed to solve this problem by having the offline shrinkwrapping process move the callback entry code into another \code{.text} ELF section while recording fixup information. The operating system can then load the callback entry code into executable pages, and the Lisp runtime can rewrite the static function pointers to point to where the code was loaded using the recorded fixup information.
